\texttt{thornado}-transport: IMEX schemes for two-moment neutrino transport respecting Fermi-Dirac statistics

### 2019-03-29 Documents/M1/Astronum_2018/sections/Numerical.tex
Workin on revisions for Astronum

In this section we consider a more ``realistic'' test: two-dimensional multigroup neutrino transport with emission, absorption, and isoenergetic scattering through a stationary background.  
This test is designed to test the realizability-preserving properties of the PD-ARS schemes.  
Figure~\ref{fig:NeutrinoStationaryTestEOS} left plots the initial thermal state: 
\begin{subequations}
\begin{align}
\text{Density: }~\rho & = 4 \times 10^{14} \times \dfrac{7.5}{( 7.5 + ( r / 5 )^4 )} ~\text{g $\cdot$ cm$^{-3}$}\\
\text{Temperature: }~T & = 1.5 \times 10^{11} \times \dfrac{1}{1 + ( r / 50 )^2 } ~\text{K}\\
\text{Electron Fraction: }~Y_e & = 0.25 \times \left(  1 + \dfrac{1}{1+ ( r / 50 )^{-12}}  \right) 
\end{align}
\label{eq:NuStatinaryStateEOS}
\end{subequations}
with radius $r$ in kilometer unit. 
Eq.~\eqref{eq:NuStatinaryStateEOS} is designed to simulate a CCSN progenitor.
Figure~\ref{fig:NeutrinoStationaryTestEOS} right plots the used opacities which are provided by interpolations of a pre-calculated opacity table based on~\cite{Bruenn_1985}.
This test is computed on Cartesian coordinate with a two-dimensional domain $D=\{\vect{x}\in\bbR^{2}:x^{1}\in[0,200], x^{2}\in[0,200]\}$ in kilometer unit, a grid of 128 elements in each direction, 10 energy groups covering $(0,300)$~MeV, a reflecting inner boundary and a homogeneous out-flowing outer boundary.
Because of the Cartesian coordinate, we ignored some geometric terms, which leads the initial condition to be non-axisymmetric.
We initialize the neutrino number density to $\cJ = 10^{-99}$, flux density to $\bcH=0$, and evolve until an approximate steady state is reached ($t=5$~ms).
No hydrodynamics is used for the stationary background.
For this test, we employed both CB and Minerbo closures.  
We attempted to run this test with our PD-ARS schemes, SSP2332 from~\cite{pareschiRusso_2005}, IMEXRKCB2 proposed by Cavaglieri \& Bewley~\cite{cavaglieriBewley2015}, and the IMEX PC2 scheme proposed by McClarren et al.~\cite{mcclarren_etal_2008}.
Only the PD-ARS schemes produce realizable moments and are able to evolve to a steady state with either CB or Minerbo closure.
SSP2332 and IMEX PC2 failed after a few time steps with either CB or Minerbo closure because of the development of unrealizable moments.
Even though IMEXRKCB2 with Minerbo closure can run and reach a steady state in this test, its results are different from that of PD-ARS schemes with CB closure, and there is no guarantee of stability.
%It's difficult to determine whether the results are in fact compromised by a particular correction step and, if so, in what way and to what degree.

Results obtained with the IMEX PD-ARS schemes are plotted in Figure~\ref{fig:NeutrinoStationaryTestEvolve} for various times: $t=0.01$~ms (top panels), 0.35~ms (middle panels), and 5.0~ms (bottom panels).  
In the left column we plot the solution in the $|\vect{\cH}|\cJ$-plane.  
In the middle column we show scatter plots of the number density $\cJ$ versus radius for select neutrino energies: 5~MeV (red lines), 16~MeV (magenta lines), and $93$~MeV (blue lines).  
In the right column we plot the flux factor $|\vect{\cH}|/\cJ$ versus radius for the same neutrino energies as in the middle column. 
In the left panels, each solution point in the domain is marked as a red dot in the $|\vect{\cH}|\cJ$-plane, and the realizable domain is shown as the light blue region.  
The figures show that all the states in the simulation with the PD-ARS schemes are realizable.
In the middle and right column, we can see how the neutrinos are generated near the core, stream out, and eventually reach an equilibrium distribution over the phase space.
The oscillation in the right column is caused by the oscillation in the numerical flux $|\vect{\cH}|$.
Parts of the causes can be the numerical error and the non-axisymmetric boundary condition.
We should also note that no flux limiter is used in this test.
The oscillation can be the spurious oscillations that occur due to discontinuities in the solution domain.
We leave these for our future study.

>>>>

In this section we consider a more ``realistic'' test: two-dimensional multigroup neutrino transport with emission, absorption, and isoenergetic scattering through a stationary background.  
\modified{This test is designed to test the realizability-preserving properties of the PD-ARS schemes.  
Figure~\ref{fig:NeutrinoStationaryTestEOS} left plots the initial thermal state: 
\begin{subequations}
\begin{align}
\text{Density: }~\rho & = 4 \times 10^{14} \times \dfrac{7.5}{( 7.5 + ( r / 5 )^4 )} ~\text{g $\cdot$ cm$^{-3}$}\\
\text{Temperature: }~T & = 1.5 \times 10^{11} \times \dfrac{1}{1 + ( r / 50 )^2 } ~\text{K}\\
\text{Electron Fraction: }~Y_e & = 0.25 \times \left(  1 + \dfrac{1}{1+ ( r / 50 )^{-12}}  \right) 
\end{align}
\label{eq:NuStatinaryStateEOS}
\end{subequations}
with radius $r$ in kilometer unit. 
Eq.~\eqref{eq:NuStatinaryStateEOS} is designed to simulate a CCSN progenitor.
Figure~\ref{fig:NeutrinoStationaryTestEOS} right plots the used opacities which are provided by interpolations of a pre-calculated opacity table based on~\cite{Bruenn_1985}.
This test is computed on Cartesian coordinate with a two-dimensional domain $D=\{\vect{x}\in\bbR^{2}:x^{1}\in[0,200], x^{2}\in[0,200]\}$ in kilometer unit, a grid of 128 elements in each direction, 10 energy groups covering $(0,300)$~MeV, a reflecting inner boundary and a homogeneous out-flowing outer boundary.
Because of the Cartesian coordinate, we ignored some geometric terms, which leads the initial condition to be non-axisymmetric.}
We initialize the neutrino number density to $\cJ = 10^{-99}$, flux density to $\bcH=0$, and evolve until an approximate steady state is reached ($t=5$~ms).
No hydrodynamics is used for the stationary background.
For this test, we employed both CB and Minerbo closures.  
We attempted to run this test with our PD-ARS schemes, SSP2332 from~\cite{pareschiRusso_2005}, IMEXRKCB2 proposed by Cavaglieri \& Bewley~\cite{cavaglieriBewley2015}, and the IMEX PC2 scheme proposed by McClarren et al.~\cite{mcclarren_etal_2008}.
Only the PD-ARS schemes produce realizable moments and are able to evolve to a steady state with either CB or Minerbo closure.
SSP2332 and IMEX PC2 failed after a few time steps with either CB or Minerbo closure because of the development of unrealizable moments.
Even though IMEXRKCB2 with Minerbo closure can run and reach a steady state in this test, its results are different from that of PD-ARS schemes with CB closure, and there is no guarantee of stability.
%It's difficult to determine whether the results are in fact compromised by a particular correction step and, if so, in what way and to what degree.

Results obtained with the IMEX PD-ARS schemes are plotted in Figure~\ref{fig:NeutrinoStationaryTestEvolve} for various times: $t=0.01$~ms (top panels), 0.35~ms (middle panels), and 5.0~ms (bottom panels).  
In the left column we plot the solution in the $|\vect{\cH}|\cJ$-plane.  
In the middle column we show scatter plots of the number density $\cJ$ versus radius for select neutrino energies: 5~MeV (red lines), 16~MeV (magenta lines), and $93$~MeV (blue lines).  
In the right column we plot the flux factor $|\vect{\cH}|/\cJ$ versus radius for the same neutrino energies as in the middle column. 
In the left panels, each solution point in the domain is marked as a red dot in the $|\vect{\cH}|\cJ$-plane, and the realizable domain is shown as the light blue region.  
The figures show that all the states in the simulation with the PD-ARS schemes are realizable.
In the middle and right column, we can see how the neutrinos are generated near the core, stream out, and eventually reach an equilibrium distribution over the phase space.
\modified{The oscillation in the right column is caused by the oscillation in the numerical flux $|\vect{\cH}|$.
Parts of the causes can be the numerical error and the non-axisymmetric boundary condition.
We should also note that no flux limiter is used in this test.
The oscillation can be the spurious oscillations that occur due to discontinuities in the solution domain.
We leave these for our future study.}

### 2019-03-29 Documents/M1/Astronum_2018/sections/TwoMoments.tex
Workin on revisions for Astronum

\subsection{Transport Equations}
We consider neutrino transport through a static background, and include neutrino-matter interactions due to emission, absorption, and isoenergetic scattering.
After scaling to dimensionless units, the Boltzmann equation can be written as
\begin{equation}
  \pd{f}{t}+\vect{\ell}\cdot\nabla f
  =\f{1}{\tau}\,\cC(f),
  \label{eq:boltzmann}
\end{equation}
where the distribution function $f = f(\omega,\varepsilon,\vect{x},t)$ gives the number of neutrinos propagating in the direction $\omega\in\bbS^{2}$, with energy $\varepsilon\in\bbR^{+}$, at position $\vect{x}\in\bbR^{3}$ and time $t\in\bbR^{+}$.  
$\vect{\ell}(\omega)\in\bbR^{3}$ is the unit vector parallel to the neutrino three-momentum: $\vect{p}=\varepsilon\,\vect{\ell}$.
On the right-hand side, $\tau$ is a collision time scale.  
In opaque regions, where neutrinos have frequent interactions with the background, $\tau\ll1$.  
In transparent regions, where neutrinos rarely interact and stream freely, $\tau\gg1$.
The collision term, $\cC(f)$, which models emission, absorption, and isoenergetic scattering is given by
\begin{equation}
  \cC(f)=\xi\,\big(\,f_{0}-f\,\big)
  +(1-\xi)\,\big(\,\f{1}{4\pi}\int_{\bbS^{2}}f\,d\omega-f\,\big),
  \label{eq:collisionTerm}
\end{equation}
where $\xi = \sigma_{\Ab} / (\sigma_{\Ab}  + \sigma_{\Scatt} )$ is the ratio of the absorption opacity $\sigma_{\Ab}$ to the total opacity.  
The scattering opacity is $\sigma_{\Scatt}$.  
The limit $\xi = 1$, when $\sigma_{\Scatt} = 0$, corresponds to pure absorption, while $\xi = 0$, when $\sigma_{\Ab} = 0$, corresponds to pure scattering.  
The equilibrium distribution function for neutrinos is given by the Fermi-Dirac distribution:
\begin{equation}
  f_{0}(\vect{z})=\f{1}{e^{(\varepsilon-\mu(\vect{x}))/T(\vect{x})}+1}, 
  \label{eq:fermiDirac}
\end{equation}
where $\vect{z}:=\{\varepsilon,\vect{x}\}$, $T$ is the material temperature in energy units and $\mu$ is the neutrino chemical potential.
Both $T$ and $\mu$ depend on the spatial position $\vect{x}$.

@@

Neutrinos are fermions and obey the Pauli exclusion principle.  
Because of this, the neutrino distribution function is bounded; i.e.~$f \in [0,1]$.
As a result, the angular moments $\cJ$ and $\vect{\cH}$ and the Eddington factor $\chi$ satisfy the following bounds~\cite{levermore_1984,lareckiBanach_2011,kershaw_1976,shohatTamarkin_1943}: 
\begin{align}
\cJ \in[0,1], \quad &(1-\cJ)\cJ-|\vect{\cH}| \geq 0, \label{eq:MomentsBounds} \\
  \chi_{\mbox{\tiny min}}
  =\max\big(1-\f{2}{3\cJ},h^{2}\big)
  \leq & \chi \leq \min\big(1,\f{1}{3\cJ}-\f{\cJ}{1-\cJ}h^{2}\big)=\chi_{\mbox{\tiny max}},
  \label{eq:eddingtonFactorBounds}
\end{align}
where $h = |\bcH|/\cJ$ is the flux factor.

>>>>

\subsection{Transport Equations}
\modified{We consider neutrino transport through a static background in a non-relativistic framework.  
For simplicity, we only include neutrino-matter interactions due to emission, absorption, and isoenergetic scattering.  
It is possible to include other interactions, such as neutrino-electron scattering and pair processes, in the two-moment model.  
However, these interactions result in dense coupling across neutrino energies and across neutrino species, and we have not yet considered the realizability-preserving property of our scheme in this case.
We leave this for future work.}

After scaling to dimensionless units, the Boltzmann equation can be written as
\begin{equation}
  \pd{f}{t}+\vect{\ell}\cdot\nabla f
  =\f{1}{\tau}\,\cC(f),
  \label{eq:boltzmann}
\end{equation}
where the distribution function $f = f(\omega,\varepsilon,\vect{x},t)$ gives the number of neutrinos propagating in the direction $\omega\in\bbS^{2}$, with energy $\varepsilon\in\bbR^{+}$, at position $\vect{x}\in\bbR^{3}$ and time $t\in\bbR^{+}$.  
$\vect{\ell}(\omega)\in\bbR^{3}$ is the unit vector parallel to the neutrino three-momentum: $\vect{p}=\varepsilon\,\vect{\ell}$.
On the right-hand side, $\tau$ is a collision time scale.  
In opaque regions, where neutrinos have frequent interactions with the background, $\tau\ll1$.  
In transparent regions, where neutrinos rarely interact and stream freely, $\tau\gg1$.
The collision term, $\cC(f)$, which models emission, absorption, and isoenergetic scattering is given by
\begin{equation}
  \cC(f)=\xi\,\big(\,f_{0}-f\,\big)
  +(1-\xi)\,\big(\,\f{1}{4\pi}\int_{\bbS^{2}}f\,d\omega-f\,\big),
  \label{eq:collisionTerm}
\end{equation}
where $\xi = \sigma_{\Ab} / (\sigma_{\Ab}  + \sigma_{\Scatt} )$ is the ratio of the absorption opacity $\sigma_{\Ab}$ to the total opacity.  
The scattering opacity is $\sigma_{\Scatt}$.  
The limit $\xi = 1$, when $\sigma_{\Scatt} = 0$, corresponds to pure absorption, while $\xi = 0$, when $\sigma_{\Ab} = 0$, corresponds to pure scattering.  
The equilibrium distribution function for neutrinos is given by the Fermi-Dirac distribution:
\begin{equation}
  f_{0}(\vect{z})=\f{1}{e^{(\varepsilon-\mu(\vect{x}))/T(\vect{x})}+1}, 
  \label{eq:fermiDirac}
\end{equation}
where $\vect{z}:=\{\varepsilon,\vect{x}\}$, $T$ is the material temperature in energy units and $\mu$ is the neutrino chemical potential.
Both $T$ and $\mu$ depend on the spatial position $\vect{x}$.

@@

Neutrinos are fermions and obey the Pauli exclusion principle.  
Because of this, the neutrino distribution function is bounded; i.e.~$f \in [0,1]$.
As a result, the angular moments $\cJ$ and $\vect{\cH}$ and the Eddington factor $\chi$ satisfy the following bounds~\cite{levermore_1984,lareckiBanach_2011,kershaw_1976,shohatTamarkin_1943}: 
\begin{align}
\cJ \in[0,1], \quad &(1-\cJ)\cJ-|\vect{\cH}| \geq 0, \label{eq:MomentsBounds} \\
  \chi_{\mbox{\tiny min}}
  =\max\big(1-\f{2}{3\cJ},h^{2}\big)
  \leq & \chi \leq \min\big(1,\f{1}{3\cJ}-\f{\cJ}{1-\cJ}h^{2}\big)=\chi_{\mbox{\tiny max}},
  \label{eq:eddingtonFactorBounds}
\end{align}
where $h = |\bcH|/\cJ$ is the flux factor.
\modified{The inequalities in Eqs.~\eqref{eq:MomentsBounds} and \eqref{eq:eddingtonFactorBounds} hold for moments as defined in Eq.~\eqref{eq:angularMoments} with $f\in[0,1]$.  
In a relativistic model (e.g., \cite{shibata_etal_2011,cardall_etal_2013}), the evolved quantities --- functions of the moments in Eq.~\eqref{eq:angularMoments} --- satisfy different bounds.  
We are currently exploring extensions of the realizability-preserving scheme presented here to the relativistic case.}
